---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tabler

<!-- badges: start -->
<!-- badges: end -->

The immediate results of statistical analyses generated by R are rarely in a form that is ideal for the intended audience's use and consumption. The `tabler` package contains functions (and accompanying strategies) we can use to make our tabular results ready for presentation and/or dissemination.

Tables are one of the most popular ways to present information to our audience. When our goal is to show our readers precise numerical summaries of our data, the exact values that resulted from our analysis are important.

However, the specific details that we may want to present from table to table, or the way in which we want to present them, can vary widely. For example, an initial table of descriptive information about our participants may or may not be stratified by subgroups of interest. Additionally, our tables may include variables of different types (i.e., numerical or categorical), and for any given type, they may include various different kinds of numerical summaries. Further, there are numerous types of inferential models that result in numerical summaries that we may want to present in a table. For these reasons, it is difficult to create a good one-size-fits-all approach to automating the process of making results ready for presentation and dissemination.

Having said that, we walk through some basic strategies for automating the processes of putting results into presentation-ready summary tables below with the help of the `tabler` package. And we attempt to do so in a way that makes updating them (which happens A LOT) as painless as possible.

## Installation

You can install the development version of `tabler` like so:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("brad-cannell/tabler")
```

## tabler functions

* **tabler:** The `tabler()` function assists with the creation of summary tables. The `tabler()` function always returns a data frame (or an error). 

## Example 1: The most basic use case

We will start by demonstrating a complete, bare minimum example without any modification to our example data or any of the default values in the `tabler()` function. After walking through this example, we will demonstrate several ways to modify the results. We will also demonstrate the effects of adjusting some of `tabler()`s default values. Finally, we will demonstrate some techniques to make the table creation process more efficient in a couple of common use cases.

Let's start by loading `tabler` and `dplyr`.

```{r example}
library(tabler)
library(dplyr, warn.conflicts = FALSE)
```

### Load data

For the purposes of making a self-contained example, the `tabler` package comes with a small example data frame that is intended to have some of the features of real study data. We will use it to demonstrate how to use `tabler` below. 

```{r}
# Load example data
data(study)
```

```{r}
glimpse(study)
```

